Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs

### 2020-02-11 manuscript/manuscript_zscore.Rmd
Ready to submit pre-print

---
title: "Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional extents, respectively. We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged z-scores (i.e. scaled anomalies) calculated on a per-station basis. Sub-regionally, temperature and chlorophyll a increased from 1993 to 2016. Total nitrogen shows a weak increase driven by low years in the early 1990s. Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll a, total nitrogen, total phosphorus, and N:P all stable over time. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. This analysis suggests an association between lake temperature and primary production. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.'
---

@@

For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on centered and scaled water quality values (i.e. z-scores) to find increasing or decreasing annual water quality trends. Details are outlined below.

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites. For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw-values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term mean over a pre-determined reference period.  This allowed all sites to be shifted to a common baseline and the deviations, or anomalies, show change over the specified reference period.

@@

Analysis that focuses on anomalies works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero µg/L to ~900 µg/L.  To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period [e.g., @jones1996calculating].  The resultant value is commonly referred to as a z-score.  We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.   Furthermore, since we are interested in water quality trends over time at individual sites (i.e. lakes), z-scores were calculated over the reference period, 1993-2016, for each site.  We refer to this approach as "site-specific z-scores".

@@

Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].  We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.  For each parameter, we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g. $\beta$~1~ = 0). Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].  Our interpretation of the trends attempts to follow this advice and we assess trends with the information provided by the magnitude of the slopes, the p-values, and our understanding of the processes involved.

### Comparison of Rhode Island to the Region

@@

Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (average z-score) in Rhode Island lakes and reservoirs. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).  Panel A. Yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.  Panel B. Yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).

@@

Mean annual trends for nutrients were weaker or showed no trend over time. The data suggest a positive trend in TN (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average z-score). Panel A. Yearly averaged TN z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).  Since 1995, the yearly trend shows a much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).  TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tp_rl$slope,3)` , p-value: `r round(ww_tp_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).  Panel A. Yearly averaged TP z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TP z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.). The pattern is the same for the TN:TP ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average z-score).  Panel A. Yearly averaged TN:TP ratio z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN:TP ratio z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).  Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from "fill in with zenodo link" [@zenodo_citation].

@@

In general, there was little evidence to suggest broad regional changes. Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP showed little change (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)

@@

Our sub-regional analysis indicates that even when nutrient regimes exhibit relative stability (i.e. neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can can occur.  Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.  This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.  Also, geographic extent does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g. northeastern and mid-western United States).  However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.

@@

Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995, @litke1999review].  However, in many lakes there are still likely sufficient nutrients present to allow for increases in chlorophyll *a*. Additionally, these results point to the fact that chlorophyll *a* and algal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of a catchment that is impervious surface and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

@@

## Data Analysis Approach

The analysis approach we used here, site-specific z-scores, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface]. However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g. @Oliver_2017], we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.  In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).  Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.

>>>>

---
title: "Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional extents, respectively. We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged z-scores (i.e., scaled anomalies) calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen shows a weak increase driven by low years in the early 1990s. Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. This analysis suggests an association between lake temperature and primary production. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.'
---

@@

For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on centered and scaled water quality values (i.e., z-scores) to find increasing or decreasing annual water quality trends. Details are outlined below.

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites. For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term mean over a pre-determined reference period.  This allowed all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.

@@

Analysis that focuses on anomalies works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero µg/L to ~900 µg/L.  To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period [e.g., @jones1996calculating].  The resultant value is commonly referred to as a z-score.  We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.   Furthermore, since we are interested in water quality trends over time at individual sites (i.e., lakes), z-scores were calculated over the reference period, 1993-2016, for each site.  We refer to this approach as "site-specific z-scores".

@@

Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].  We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.  For each parameter, we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0). Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].  Our interpretation of the trends attempts to follow this advice and we assess trends with the information provided by the magnitude of the slopes, the p-values, and our understanding of the processes involved.

### Comparison of Rhode Island to the region

@@

Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (average z-score) in Rhode Island lakes and reservoirs. Points are averages of site-specific z-scores and ranges are standard deviations of site-specific z-scores. Blue indicates yearly site-specific z-scores that were, on average, below the site-specific long-term means. Red indicates yearly site-specific z-scores that were, on average, above the site-specific long-term means.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (average z-score).  Panel A. Yearly averaged chlorophyll a z-scores from the URI Watershed Watch data.  Panel B. Yearly averaged chlorophyll a z-scores from the LAGOSNE dataset. Points are averages of site-specific z-scores and ranges are standard deviations of site-specific z-scores. Blue indicates yearly site-specific z-scores that were, on average, below the site-specific long-term means. Red indicates yearly site-specific z-scores that were, on average, above the site-specific long-term means.", display = "cite")`A.).

@@

Mean annual trends for nutrients were weaker or showed no trend over time. The data suggest a positive trend in TN (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average z-score). Panel A. Yearly averaged TN z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN z-scores from the LAGOSNE dataset. Points are averages of site-specific z-scores and ranges are standard deviations of site-specific z-scores. Blue indicates yearly site-specific z-scores that were, on average, below the site-specific long-term means. Red indicates yearly site-specific z-scores that were, on average, above the site-specific long-term means.", display = "cite")`A.).  Since 1995, the yearly trend shows a much lower increase over time (slope = `r signif(ww_tn_rl_no9394$slope,2)`, p = `r signif(ww_tn_rl_no9394$p.value, 2)`).  TP does not show a trend over time in the yearly z-scores (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).  Panel A. Yearly averaged TP z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TP z-scores from the LAGOSNE dataset. Points are averages of site-specific z-scores and ranges are standard deviations of site-specific z-scores. Blue indicates yearly site-specific z-scores that were, on average, below the site-specific long-term means. Red indicates yearly site-specific z-scores that were, on average, above the site-specific long-term means.", display = "cite")`A.). The pattern is the same for the TN:TP ratio (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average z-score).  Panel A. Yearly averaged TN:TP ratio z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN:TP ratio z-scores from the LAGOSNE dataset. Points are averages of site-specific z-scores and ranges are standard deviations of site-specific z-scores. Blue indicates yearly site-specific z-scores that were, on average, below the site-specific long-term means. Red indicates yearly site-specific z-scores that were, on average, above the site-specific long-term means.", display = "cite")`A.).  Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from "fill in with zenodo link" [@zenodo_citation].

@@

In general, there was little evidence to suggest broad regional changes. Chlorophyll *a* showed a very weak negative trend (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B.)

@@

Our sub-regional analysis indicates that even when nutrient regimes exhibit relative stability (i.e., neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can occur.  Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.  This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.  Also, geographic extent does indeed matter when trying to identify long-term water quality trends. Similar to the results of Oliver et al. [-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g., northeastern and mid-western United States).  However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.

@@

Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995; @litke1999review].  However, in many lakes there are still likely sufficient nutrients present to allow for increases in chlorophyll *a*. Additionally, these results point to the fact that chlorophyll *a* and algal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of a catchment that is impervious surface and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

@@

## Data analysis approach

The analysis approach we used here, site-specific z-scores, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface]. However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g., @Oliver_2017], we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.  In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).  Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.
